---
title: "Movement quality prediction"
author: "EG"
date: "January 29, 2016"
output: 
  html_document: 
    keep_md: yes
---
## Overview
In this report, we aim to build a predictive model of the quality of movement recorded by fitness devices. In particular, we build a random forest model predicting movement classification based on accelleration measurements on the belt, forearm, dumbbell, and arm. There are five movement classes (from http://groupware.les.inf.puc-rio.br/har):

1) A: movement exactly according to specification (correct form)

2) B: throwing elbows to front (mistake)

3) C: lifting dumbbell only half way (mistake)

4) D: lowering dumbell only half way (mistake)

5) E: throwing hips to the front (mistake)

We find that a random forest model with 10-fold cross-validation generates a lower out of sample error than a linear discriminant analysis with 10-fold cross-validation trained on the same dataset (4.5% error vs. 48.2% error, respectively). Our predictive algorithm correctly predicts 19/20 of the classes in the test set.

## Pre-processing
Load libraries necessary for analysis.
``` {r}
library(caret)
library(randomForest)
library(doParallel)
```

Load in training and testing sets.
``` {r}
url1 <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
download.file(url1, "training.csv", method = "curl")

url2 <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
download.file(url2, "testing.csv", method = "curl")

training <- read.table("training.csv", sep = ",", header = T, as.is = T)
testing <- read.table("testing.csv", sep = ",", header = T, as.is = T)
```

Set seed for analyses.
```{r}
set.seed(2222)
```

Limit variables to those relating to accelaration on belt, forearm, arm, and dumbbell. Also include the outcome variable.
``` {r}
accell <- grep("accel", colnames(training))
training_accel <- training[,c(accell, which(colnames(training) == "classe"))]
```

Determine whethere there are any variables with missing values that can be removed from the dataset. Variance of accelleration in all areas missing the majority of observations; remove those variables.
``` {r}
summary(training_accel)

varAccell <- grep("^var", colnames(training_accel))
training_accel <- training_accel[,-varAccell]
```

Convert the outcome variables to factors to prepare for model building.
``` {r}
training_accel$classe = as.factor(training_accel$classe)
```

## Predictive model
Use random forest algorithm to model excercise data, due to its high accuracy. Due to high computational load, allow for parallel processing to speed the process.

Specify that 10-fold cross-validation should be used, include control parameters in train function.
``` {r}
cluster <- makeCluster(detectCores() - 1)
registerDoParallel(cluster)

fitControl <- trainControl(method = "cv",
                           number = 10,
                           allowParallel = T)

modFit <- train(classe ~., training_accel, method = "rf", trControl = fitControl)

stopCluster(cluster)
```

Based on the 10-fold cross-validation, evaluate the out of sample error. Accuracy across the folds is 95.5%, with 4.5% estimated out of sample error.
``` {r}
confusionMatrix(modFit)
accuracy <- 27.6 + 18 + 16.7 + 15.3 + 17.9
error <- 100 - accuracy
error
```

### Predictions
Match testing set to training set.
``` {r}
accell_test <- grep("accel", colnames(testing))
testing_accel <- testing[,c(accell_test, which(colnames(testing) == "problem_id"))]
varAccell_test <- grep("^var", colnames(testing_accel))
testing_accel <- testing_accel[,-varAccell_test]
```

Use random forest model to predict movement classification in test set.
``` {r}
pred <- predict(modFit, testing_accel)
testing_accel$classe <- pred
testing_accel[, 17:18]
```

## Supplemental analysis
Use linear discriminant analysis for the same set of data to confirm random forest is a good choice for the analysis.
``` {r}
cluster <- makeCluster(detectCores() - 1)
registerDoParallel(cluster)

fitControl <- trainControl(method = "cv",
                           number = 10,
                           allowParallel = T)

modFit2 <- train(classe ~., training_accel, method = "lda", trControl = fitControl)

stopCluster(cluster)
```

Estimate out of sample error from cross-validation. Accuracy is 51.7%, error is 48.3%. This confirms that random forest is a good choice for this analysis.
``` {r}
confusionMatrix(modFit2)
accuracy <- 19.6 + 8.2 + 5.4 + 10.3 + 8.2
error2 = 100 - accuracy
error2
```